---
title: "Project1 Paper"
author: "Likhitha, Paulina, Shrihan"
date: "2023-10-24"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
library(ezids)
library(ggplot2)
#install.packages("tigris")
library(tigris)
#install.packages("sf")
library(sf)
#install.packages("here")
library(here)
#install.packages("dplyr")
library(dplyr)
#install.packages("leaflet")
library(leaflet)
#install.packages("reshape2")
library(reshape2)
#install.packages("tmap")
library(tmap)
#install.packages("viridis")
library(viridis)
#install.packages("stargazer")
library(stargazer)
```

## Introduction and SVI

Text about the Introduction

## Data set and Variables

Text and code showing the Dataset and variables

## Clean the Data

Load the data file

```{r Load the data file}
SVI_Data <- read.csv("SVI_2020_US.csv")
head(SVI_Data)
```

Keep relevant columns

```{r select columns}

#selecting the required columns by subset function

Clean_data <- subset(SVI_Data, select = c(ST,STATE,ST_ABBR,STCNTY,COUNTY,FIPS,LOCATION,AREA_SQMI,EPL_POV150,	EPL_UNEMP,	EPL_HBURD,	EPL_NOHSDP,	EPL_UNINSUR,	SPL_THEME1,	RPL_THEME1,	EPL_AGE65,	EPL_AGE17,	EPL_DISABL,	EPL_SNGPNT,	EPL_LIMENG,	SPL_THEME2,	RPL_THEME2,	EPL_MINRTY,	SPL_THEME3,	RPL_THEME3, E_MINRTY, EP_HISP, EP_ASIAN, EP_AIAN, EPL_MUNIT,	EPL_MOBILE,	EPL_CROWD,	EPL_NOVEH,	EPL_GROUPQ,	SPL_THEME4,	RPL_THEME4,	SPL_THEMES,	RPL_THEMES, E_AGE65, EP_POV150, EP_AGE65, EP_NOHSDP
) )

head(Clean_data)


```

subset Clean_Data to just CA tracts

```{r CA tracts}
CA_SVI <- subset(Clean_data, ST_ABBR == "CA")
```

Show Outliers

```{r Outliers}
#Display the Outliers in the cleaned data

Outliers1 = outlierKD2(CA_SVI, RPL_THEME1, rm = FALSE, qqplt = TRUE)
Outliers2 = outlierKD2(CA_SVI, RPL_THEME2, rm = FALSE, qqplt = TRUE)
Outliers3 = outlierKD2(CA_SVI, RPL_THEME3, rm = FALSE, qqplt = TRUE)
Outliers4 = outlierKD2(CA_SVI, RPL_THEME4, rm = FALSE, qqplt = TRUE)
Outliers = outlierKD2(CA_SVI, RPL_THEMES, rm = FALSE, qqplt = TRUE)

```

## Data Analysis

Code and findings for each of Maps, Scatterplots and histograms

## Correlation Heatmap

## Hypothesis Testing

## Conclusion

## Future Work
