---
title: "EDA"
author: "Likhitha, Paulina, Shrihan"
date: "2023-10-10"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---


```{r, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(scientific=T, digits = 3) 
# Restart R session

# Load necessary packages
library(ezids)
library(tigris)
library(sf)
library(here)
library(dplyr)
library(leaflet)
library(reshape2)
library(tmap)
library(viridis)
library(stargazer)
library(maps)
library(ggplot2)

```

**First Look at Dataset**

```{r}

SVI_Data <- read.csv("SVI_2020_US.csv")
us_cases <- read.csv('USAFacts/confirmed-covid-19-cases-in-us-by-state-and-county.csv')
us_deaths <- read.csv('USAFacts/confirmed-covid-19-deaths-in-us-by-state-and-county.csv')

head(SVI_Data)
head(us_cases)
head(us_deaths)

```

**Column Subsetting and Renaming**


```{r}

SVI <- subset(SVI_Data, select = c(STATE,STCNTY,COUNTY,FIPS,EP_POV150,EP_LIMENG,EP_AGE65, EP_MINRTY,	EP_UNEMP,	EP_NOHSDP,EP_AGE17,EP_DISABL,EP_SNGPNT,EP_MUNIT,EP_MOBILE,EP_CROWD,EP_NOVEH,EP_GROUPQ,RPL_THEMES) )

#SVI1 <- subset(SVI_Data , select=c(EPL_UNEMP,EPL_HBURD,EPL_NOHSDP,EPL_UNINSUR,SPL_THEME1,RPL_THEME1))
SVI2 <- subset(SVI_Data , select=c(STATE,STCNTY,COUNTY,FIPS,EPL_UNEMP,EPL_HBURD,EPL_NOHSDP,EPL_UNINSUR,SPL_THEME1,RPL_THEME1,EPL_AGE65,EPL_AGE17,EPL_DISABL,EPL_SNGPNT,EPL_LIMENG,SPL_THEME2,RPL_THEME2,EPL_MINRTY,SPL_THEME3,RPL_THEME3,EPL_MUNIT,EPL_MOBILE,EPL_CROWD,EPL_NOVEH,EPL_GROUPQ,SPL_THEME4,RPL_THEME4))
#SVI3 <- subset(SVI_Data , select=c(EPL_MINRTY,SPL_THEME3,RPL_THEME3))
#SVI4 <- subset(SVI_Data , select=c(EPL_MUNIT,EPL_MOBILE,EPL_CROWD,EPL_NOVEH,EPL_GROUPQ,SPL_THEME4,RPL_THEME4))

# Renaming columns
colnames(SVI)[colnames(SVI) == 'STCNTY'] <- 'county_fips'
colnames(SVI2)[colnames(SVI2) == 'STCNTY'] <- 'county_fips'

us_cases <- rename(us_cases, confirmed_cases = confirmed)
head(SVI2)


```



**Missing values removal**
```{r}
# Filter out rows with any value equal to -999
SVI <- SVI[rowSums(SVI == -999, na.rm = TRUE) == 0, ]
SVI2 <- SVI2[rowSums(SVI2 == -999, na.rm = TRUE) == 0, ]


# Drop rows with missing values
SVI <- na.omit(SVI)

SVI2 <- na.omit(SVI2)



us_cases[is.na(us_cases$confirmed_cases), ]
us_deaths[is.na(us_deaths$deaths), ]

# Display summary statistics for each column
head(SVI)
head(SVI2)


```

**Row Subsetting**
```{r}
CA_SVI <- subset(SVI, STATE == "California")
head(CA_SVI)

```

**Formatting**

```{r}
library(dplyr)

format_str <- '%Y-%m-%d'

# Convert the 'date' column to datetime and create a new 'datetime' column
us_cases$datetime <- as.POSIXct(us_cases$date, format = format_str)
us_deaths$datetime <- as.POSIXct(us_deaths$date, format = format_str)


# Drop rows that do not refer to a specific county ('Statewide Unallocated')
us_cases <- us_cases[us_cases$county_name != 'Statewide Unallocated', ]
us_deaths <- us_deaths[us_deaths$county_name != 'Statewide Unallocated', ]


us_cases_ca <- subset(us_cases, state_name == "CA")
us_deaths_ca <- subset(us_deaths, state_name == "CA")


# Display the first few rows of the resulting data frame
head(us_cases_ca)
head(us_deaths_ca)

```


```{r}
# Calculate Sum of confirmed cases based on county fips and groupby county 
us_cases_grouping_fips <- us_cases_ca %>%
  group_by(county_fips, state_name) %>%
  summarise(max_confirmed_cases = sum(confirmed_cases)) %>%
  arrange(desc(max_confirmed_cases))

us_cases_grouping_county <- us_cases_ca %>%
  group_by(county_name, state_name) %>%
  summarise(max_confirmed_cases = sum(confirmed_cases)) %>%
  arrange(desc(max_confirmed_cases))

result <- bind_cols(us_cases_grouping_fips, us_cases_grouping_county)
Summarized_cases<-result[, 1:4]
new_col_names <- c("fips", "state", "confirmed", "county")
names(Summarized_cases) <- new_col_names

# Calculate Sum of deaths based on county fips and groupby county 

us_deaths_grouping_fips <- us_deaths_ca %>%
  group_by(county_fips, state_name) %>%
  summarise(max_deaths = sum(deaths)) %>%
  arrange(desc(max_deaths))

us_deaths_grouping_county <- us_deaths_ca %>%
  group_by(county_name, state_name) %>%
  summarise(max_deaths = sum(deaths)) %>%
  arrange(desc(max_deaths))

result1 <- bind_cols(us_deaths_grouping_fips, us_deaths_grouping_county)
Summarized_deaths<-result1[, 1:4]
new_col_names1 <- c("fips", "state", "deaths", "county")
names(Summarized_deaths) <- new_col_names1

```

**Map of CA colored by confirmed cases**

```{r}

# Install and load the usmap package
library(usmap)
library(plotly)


# Plot the choropleth map
plot_usmap(
  data = Summarized_cases,
  values = "confirmed",
  include = c("CA"),
  labels = TRUE
) + 
scale_fill_gradient(
  low = "lightblue", high = "darkblue", name = "Confirmed Cases", label = scales::comma) + 
labs(title = "COVID-19 Confirmed Cases in California Counties") +
theme(legend.position = "right")

```


**Map of CA colored by deaths**

```{r}

# Install and load the usmap package
library(usmap)
library(plotly)


# Plot the choropleth map
plot_usmap(
  data = Summarized_deaths,
  values = "deaths",
  include = c("CA"),
  labels = TRUE
) + 
scale_fill_gradient(
  low = "lightblue", high = "darkblue", name = "Number of Deaths", label = scales::comma) + 
labs(title = "COVID-19 Deaths in California Counties") +
theme(legend.position = "right")

```

```{r}
us_cases_all_counties <- us_cases_ca %>%
  group_by(county_fips, county_name, state_name) %>%
  summarise(confirmed_cases = sum(confirmed_cases)) %>%
  arrange(desc(confirmed_cases))

# Group by Fips, County, and State, and calculate the mean of selected indicators
svi_all_counties <- SVI %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EP_POV150 = mean(EP_POV150),
    EP_LIMENG = mean(EP_LIMENG),
    EP_AGE65 = mean(EP_AGE65),
    EP_MINRTY = mean(EP_MINRTY),
    EP_UNEMP = mean(EP_UNEMP),
    EP_NOHSDP = mean(EP_NOHSDP),
    EP_AGE17 = mean(EP_AGE17),
    EP_DISABL = mean(EP_DISABL),
    EP_SNGPNT = mean(EP_SNGPNT),
    EP_MUNIT = mean(EP_MUNIT),
    EP_MOBILE = mean(EP_MOBILE),
    EP_CROWD = mean(EP_CROWD),
    EP_NOVEH = mean(EP_NOVEH),
    EP_GROUPQ = mean(EP_GROUPQ),
    RPL_THEMES = mean(RPL_THEMES)
  )
svi_all_counties_ca <- subset(svi_all_counties, STATE == "California")
# Merge with us_cases_all_counties and sort by confirmed_cases
svi_all_counties_cases <- merge(svi_all_counties_ca, us_cases_all_counties, by = 'county_fips') %>%
  arrange(desc(confirmed_cases))

# Remove unnecessary columns
svi_all_counties_cases <- select(svi_all_counties_cases, -county_name, -state_name)

# Display the result
head(svi_all_counties_cases)

```


```{r}

us_deaths_all_counties <- us_deaths_ca %>%
  group_by(county_fips, county_name, state_name) %>%
  summarise(deaths = sum(deaths)) %>%
  arrange(desc(deaths))

# Group by Fips, County, and State, and calculate the mean of selected indicators
svi1_all_counties <- SVI %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EP_POV150 = mean(EP_POV150),
    EP_LIMENG = mean(EP_LIMENG),
    EP_AGE65 = mean(EP_AGE65),
    EP_MINRTY = mean(EP_MINRTY),
    EP_UNEMP = mean(EP_UNEMP),
    EP_NOHSDP = mean(EP_NOHSDP),
    EP_AGE17 = mean(EP_AGE17),
    EP_DISABL = mean(EP_DISABL),
    EP_SNGPNT = mean(EP_SNGPNT),
    EP_MUNIT = mean(EP_MUNIT),
    EP_MOBILE = mean(EP_MOBILE),
    EP_CROWD = mean(EP_CROWD),
    EP_NOVEH = mean(EP_NOVEH),
    EP_GROUPQ = mean(EP_GROUPQ),
    RPL_THEMES = mean(RPL_THEMES)
  )
svi1_all_counties_ca <- subset(svi1_all_counties, STATE == "California")
# Merge with us_cases_all_counties and sort by confirmed_cases
svi_all_counties_deaths <- merge(svi1_all_counties_ca, us_deaths_all_counties, by = 'county_fips') %>%
  arrange(desc(deaths))

# Remove unnecessary columns
svi_all_counties_deaths <- select(svi_all_counties_deaths, -county_name, -state_name)

# Display the result
head(svi_all_counties_deaths)










```













```{r}


# Group by Fips, County, and State, and calculate the mean of selected indicators
SVI2_all_counties <- SVI2 %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EPL_UNEMP = mean(EPL_UNEMP),
    EPL_HBURD = mean(EPL_HBURD),
    EPL_NOHSDP = mean(EPL_NOHSDP),
    EPL_UNINSUR = mean(EPL_UNINSUR),
    SPL_THEME1 = mean(SPL_THEME1),
    RPL_THEME1 = mean(RPL_THEME1),
    EPL_AGE65 = mean(EPL_AGE65),
    EPL_AGE17 = mean(EPL_AGE17),
    EPL_DISABL = mean(EPL_DISABL),
    EPL_SNGPNT = mean(EPL_SNGPNT),
    EPL_LIMENG = mean(EPL_LIMENG),
    SPL_THEME2 = mean(SPL_THEME2),
    RPL_THEME2 = mean(RPL_THEME2),
    EPL_MINRTY = mean(EPL_MINRTY),
    SPL_THEME3 = mean(SPL_THEME3),
    RPL_THEME3 = mean(RPL_THEME3),
    EPL_MUNIT = mean(EPL_MUNIT),
    EPL_MOBILE = mean(EPL_MOBILE),
    EPL_CROWD = mean(EPL_CROWD),
    EPL_NOVEH = mean(EPL_NOVEH),
    EPL_GROUPQ = mean(EPL_GROUPQ),
    SPL_THEME4 = mean(SPL_THEME4),
    RPL_THEME4 = mean(RPL_THEME4)
  )

SVI2_all_counties_ca <- subset(SVI2_all_counties, STATE == "California")

# Merge with us_cases_all_counties and sort by confirmed_cases
SVI2_all_counties_cases <- merge(SVI2_all_counties_ca, us_cases_all_counties, by = 'county_fips') %>%
  arrange(desc(confirmed_cases))

# Remove unnecessary columns
SVI2_all_counties_cases <- select(SVI2_all_counties_cases, -county_name, -state_name)

# Display the result
head(SVI2_all_counties_cases)


```




```{r}


# Group by Fips, County, and State, and calculate the mean of selected indicators
SVI3_all_counties <- SVI2 %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EPL_UNEMP = mean(EPL_UNEMP),
    EPL_HBURD = mean(EPL_HBURD),
    EPL_NOHSDP = mean(EPL_NOHSDP),
    EPL_UNINSUR = mean(EPL_UNINSUR),
    SPL_THEME1 = mean(SPL_THEME1),
    RPL_THEME1 = mean(RPL_THEME1),
    EPL_AGE65 = mean(EPL_AGE65),
    EPL_AGE17 = mean(EPL_AGE17),
    EPL_DISABL = mean(EPL_DISABL),
    EPL_SNGPNT = mean(EPL_SNGPNT),
    EPL_LIMENG = mean(EPL_LIMENG),
    SPL_THEME2 = mean(SPL_THEME2),
    RPL_THEME2 = mean(RPL_THEME2),
    EPL_MINRTY = mean(EPL_MINRTY),
    SPL_THEME3 = mean(SPL_THEME3),
    RPL_THEME3 = mean(RPL_THEME3),
    EPL_MUNIT = mean(EPL_MUNIT),
    EPL_MOBILE = mean(EPL_MOBILE),
    EPL_CROWD = mean(EPL_CROWD),
    EPL_NOVEH = mean(EPL_NOVEH),
    EPL_GROUPQ = mean(EPL_GROUPQ),
    SPL_THEME4 = mean(SPL_THEME4),
    RPL_THEME4 = mean(RPL_THEME4)
  )

SVI3_all_counties_ca <- subset(SVI3_all_counties, STATE == "California")

# Merge with us_cases_all_counties and sort by confirmed_cases
SVI2_all_counties_deaths <- merge(SVI3_all_counties_ca, us_deaths_all_counties, by = 'county_fips') %>%
  arrange(desc(deaths))

# Remove unnecessary columns
SVI2_all_counties_deaths <- select(SVI2_all_counties_deaths, -county_name, -state_name)

# Display the result
head(SVI2_all_counties_deaths)


```








**Corelation matrix between confirmed cases and all variables**
```{r}
# Plot the correlation matrix
library(corrplot)
# Select relevant columns
svi_all_counties_cases_corr <- svi_all_counties_cases[c('confirmed_cases', 'EP_POV150', 'EP_LIMENG', 'EP_AGE65', 'EP_MINRTY', 'EP_UNEMP', 'EP_NOHSDP', 'EP_AGE17', 'EP_DISABL', 'EP_SNGPNT', 'EP_MUNIT', 'EP_MOBILE', 'EP_CROWD', 'EP_NOVEH', 'EP_GROUPQ', 'RPL_THEMES')]

# Calculate correlation matrix
corr <- cor(svi_all_counties_cases_corr)
corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.5,mar = c(0, 0, 2, 0), tl.cex = 0.5, tl.col = 'black', tl.srt = 45)



```



**Corelation matrix between deaths and all variables**
```{r}
# Plot the correlation matrix
library(corrplot)
# Select relevant columns
svi_all_counties_deaths_corr <- svi_all_counties_deaths[c('deaths', 'EP_POV150', 'EP_LIMENG', 'EP_AGE65', 'EP_MINRTY', 'EP_UNEMP', 'EP_NOHSDP', 'EP_AGE17', 'EP_DISABL', 'EP_SNGPNT', 'EP_MUNIT', 'EP_MOBILE', 'EP_CROWD', 'EP_NOVEH', 'EP_GROUPQ', 'RPL_THEMES')]

# Calculate correlation matrix
corr <- cor(svi_all_counties_deaths_corr)
corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.5,mar = c(0, 0, 2, 0), tl.cex = 0.5, tl.col = 'black', tl.srt = 45)



```

**Correlation matrix with different themes **
*Positive Correlations:*

*There are positive correlations between EPL_UNEMP and several variables, such as SPL_THEME1, RPL_THEME1, EPL_HBURD, EPL_NOHSDP, and EPL_UNINSUR. This suggests that as unemployment increases, these variables tend to increase as well.*

*There are positive correlations between EPL_MOBILE and EPL_DISABL, EPL_SNGPNT, and EPL_MINRTY. This implies that areas with higher mobile home occupancy tend to have higher percentages of disabled individuals, single-parent households, and minority populations.*

*Negative Correlations:*

*There is a negative correlation between EPL_AGE65 and several variables, such as EPL_UNEMP, EPL_MOBILE, EPL_DISABL, and EPL_MUNIT. This suggests that areas with a higher percentage of elderly population tend to have lower unemployment rates, fewer mobile homes, fewer disabled individuals, and fewer multi-unit structures.*

*There is a negative correlation between EPL_MUNIT and EPL_NOHSDP, indicating that areas with more multi-unit structures tend to have lower percentages of households with no high school diploma.*

```{r}
# Plot the correlation matrix
library(corrplot)
# Select relevant columns
SVI2_all_counties_cases_corr <- SVI2_all_counties_cases[c('confirmed_cases','EPL_UNEMP','EPL_HBURD','EPL_NOHSDP','EPL_UNINSUR','SPL_THEME1','RPL_THEME1','EPL_AGE65','EPL_AGE17','EPL_DISABL','EPL_SNGPNT','EPL_LIMENG','SPL_THEME2','RPL_THEME2','EPL_MINRTY','SPL_THEME3','RPL_THEME3','EPL_MUNIT','EPL_MOBILE','EPL_CROWD','EPL_NOVEH','EPL_GROUPQ','SPL_THEME4','RPL_THEME4')]

# Calculate correlation matrix
corr <- cor(SVI2_all_counties_cases_corr)
corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.4,mar = c(0, 0, 2, 0), tl.cex = 0.4, tl.col = 'black', tl.srt = 45)

xkabledply(corr)
```


**Correlation matrix with different themes**
```{r}
# Plot the correlation matrix
library(corrplot)
# Select relevant columns
SVI2_all_counties_deaths_corr <- SVI2_all_counties_deaths[c('deaths','EPL_UNEMP','EPL_HBURD','EPL_NOHSDP','EPL_UNINSUR','SPL_THEME1','RPL_THEME1','EPL_AGE65','EPL_AGE17','EPL_DISABL','EPL_SNGPNT','EPL_LIMENG','SPL_THEME2','RPL_THEME2','EPL_MINRTY','SPL_THEME3','RPL_THEME3','EPL_MUNIT','EPL_MOBILE','EPL_CROWD','EPL_NOVEH','EPL_GROUPQ','SPL_THEME4','RPL_THEME4')]

# Calculate correlation matrix
corr <- cor(SVI2_all_counties_deaths_corr)
corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.4,mar = c(0, 0, 2, 0), tl.cex = 0.4, tl.col = 'black', tl.srt = 45)

xkabledply(corr)
```


**Confirmed cases**

```{r}

# Which are the 10 most affected US counties?
ca_cases_top_10 <- head(us_cases_all_counties, 10)
print(ca_cases_top_10)

# Preparing a list with the 10 most affected counties:
top_10_cases_fips_list <- ca_cases_top_10$county_fips

ca_cases_top_10_datetime <- subset(us_cases_ca, county_fips %in% top_10_cases_fips_list)
head(ca_cases_top_10_datetime)



```


**Deaths**
```{r}

# Which are the 10 most affected US counties?
ca_deaths_top_10 <- head(us_deaths_all_counties, 10)
print(ca_cases_top_10)

# Preparing a list with the 10 most affected counties:
top_10_deaths_fips_list <- ca_deaths_top_10$county_fips

ca_deaths_top_10_datetime <- subset(us_deaths_ca, county_fips %in% top_10_deaths_fips_list)
head(ca_deaths_top_10_datetime)

```





**Create timeline of covid confirmed cases**
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Plot Timeline of cases
ca_cases_top_10_datetime$datetime <- as.Date(ca_cases_top_10_datetime$datetime)  # Make sure datetime is in Date format
head(ca_cases_top_10_datetime)

f <- ggplot(ca_cases_top_10_datetime, aes(x = datetime, y = confirmed_cases, color = county_name)) +
  geom_line() +
  labs(x = 'Timeline', y = 'Confirmed Cases', title = 'Confirmed cases in CA counties') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_discrete(name = 'County Name')+
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y")   # Format the x-axis with month and year


print(f)




```

**Create timeline of covid death**
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Plot Timeline of cases
ca_deaths_top_10_datetime$datetime <- as.Date(ca_deaths_top_10_datetime$datetime)  # Make sure datetime is in Date format
head(ca_deaths_top_10_datetime)

f <- ggplot(ca_deaths_top_10_datetime, aes(x = datetime, y = deaths, color = county_name)) +
  geom_line() +
  labs(x = 'Timeline', y = 'deaths', title = 'deaths in CA counties') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_discrete(name = 'County Name')+
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y")   # Format the x-axis with month and year


print(f)




```


**Create correlation matrix for just the most affected counties**
```{r}

# Select the rows in the SVI dataframe that contain the ten counties with most confirmed cases.
svi_top_10_counties_cases <- SVI[SVI$county_fips %in% top_10_cases_fips_list, ]

# Group by county, COUNTY, and STATE, and calculate the mean values for selected indicators.
svi_top_10_counties_cases <- svi_top_10_counties_cases %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EP_POV150 = mean(EP_POV150),
    EP_LIMENG = mean(EP_LIMENG),
    EP_AGE65 = mean(EP_AGE65),
    EP_MINRTY = mean(EP_MINRTY),
    EP_UNEMP = mean(EP_UNEMP),
    EP_NOHSDP = mean(EP_NOHSDP),
    EP_AGE17 = mean(EP_AGE17),
    EP_DISABL = mean(EP_DISABL),
    EP_SNGPNT = mean(EP_SNGPNT),
    EP_MUNIT = mean(EP_MUNIT),
    EP_MOBILE = mean(EP_MOBILE),
    EP_CROWD = mean(EP_CROWD),
    EP_NOVEH = mean(EP_NOVEH),
    EP_GROUPQ = mean(EP_GROUPQ),
    RPL_THEMES = mean(RPL_THEMES)
  ) %>%
  ungroup()

# Merge county SVI with confirmed cases of Covid-19
svi_top_10_counties_cases <- merge(svi_top_10_counties_cases, ca_cases_top_10, by = 'county_fips') %>%
  arrange(desc(confirmed_cases)) %>%
  select(-county_name, -state_name)

# Select columns for correlation analysis
svi_top_10_counties_cases_corr <- svi_top_10_counties_cases[c(
  'confirmed_cases', 'EP_POV150', 'EP_LIMENG', 'EP_AGE65', 'EP_MINRTY',
  'EP_UNEMP', 'EP_NOHSDP', 'EP_AGE17', 'EP_DISABL', 'EP_SNGPNT',
  'EP_MUNIT', 'EP_MOBILE', 'EP_CROWD', 'EP_NOVEH', 'EP_GROUPQ', 'RPL_THEMES'
)]

# Calculate correlation matrix
corr <- cor(svi_top_10_counties_cases_corr)

corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.5,mar = c(0, 0, 2, 0), tl.cex = 0.5, tl.col = 'black', tl.srt = 45)



```







**Create correlation matrix for just the most affected counties for deaths**
```{r}

# Select the rows in the SVI dataframe that contain the ten counties with most confirmed cases.
svi_top_10_counties_deaths <- SVI[SVI$county_fips %in% top_10_deaths_fips_list, ]

# Group by county, COUNTY, and STATE, and calculate the mean values for selected indicators.
svi_top_10_counties_deaths <- svi_top_10_counties_deaths %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EP_POV150 = mean(EP_POV150),
    EP_LIMENG = mean(EP_LIMENG),
    EP_AGE65 = mean(EP_AGE65),
    EP_MINRTY = mean(EP_MINRTY),
    EP_UNEMP = mean(EP_UNEMP),
    EP_NOHSDP = mean(EP_NOHSDP),
    EP_AGE17 = mean(EP_AGE17),
    EP_DISABL = mean(EP_DISABL),
    EP_SNGPNT = mean(EP_SNGPNT),
    EP_MUNIT = mean(EP_MUNIT),
    EP_MOBILE = mean(EP_MOBILE),
    EP_CROWD = mean(EP_CROWD),
    EP_NOVEH = mean(EP_NOVEH),
    EP_GROUPQ = mean(EP_GROUPQ),
    RPL_THEMES = mean(RPL_THEMES)
  ) %>%
  ungroup()

# Merge county SVI with confirmed cases of Covid-19
svi_top_10_counties_deaths <- merge(svi_top_10_counties_deaths, ca_deaths_top_10, by = 'county_fips') %>%
  arrange(desc(deaths)) %>%
  select(-county_name, -state_name)

# Select columns for correlation analysis
svi_top_10_counties_deaths_corr <- svi_top_10_counties_deaths[c(
  'deaths', 'EP_POV150', 'EP_LIMENG', 'EP_AGE65', 'EP_MINRTY',
  'EP_UNEMP', 'EP_NOHSDP', 'EP_AGE17', 'EP_DISABL', 'EP_SNGPNT',
  'EP_MUNIT', 'EP_MOBILE', 'EP_CROWD', 'EP_NOVEH', 'EP_GROUPQ', 'RPL_THEMES'
)]

# Calculate correlation matrix
corr <- cor(svi_top_10_counties_deaths_corr)

corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.5,mar = c(0, 0, 2, 0), tl.cex = 0.5, tl.col = 'black', tl.srt = 45)



```




**Correlation matrix for ten counties with most confirmed cases along different themes**

```{r}

# Select the rows in the SVI dataframe that contain the ten counties with most confirmed cases.
SVI2_top_10_counties_cases <- SVI2[SVI2$county_fips %in% top_10_cases_fips_list, ]

# Group by county, COUNTY, and STATE, and calculate the mean values for selected indicators.
SVI2_top_10_counties_cases <- SVI2_top_10_counties_cases %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EPL_UNEMP = mean(EPL_UNEMP),
    EPL_HBURD = mean(EPL_HBURD),
    EPL_NOHSDP = mean(EPL_NOHSDP),
    EPL_UNINSUR = mean(EPL_UNINSUR),
    SPL_THEME1 = mean(SPL_THEME1),
    RPL_THEME1 = mean(RPL_THEME1),
    EPL_AGE65 = mean(EPL_AGE65),
    EPL_AGE17 = mean(EPL_AGE17),
    EPL_DISABL = mean(EPL_DISABL),
    EPL_SNGPNT = mean(EPL_SNGPNT),
    EPL_LIMENG = mean(EPL_LIMENG),
    SPL_THEME2 = mean(SPL_THEME2),
    RPL_THEME2 = mean(RPL_THEME2),
    EPL_MINRTY = mean(EPL_MINRTY),
    SPL_THEME3 = mean(SPL_THEME3),
    RPL_THEME3 = mean(RPL_THEME3),
    EPL_MUNIT = mean(EPL_MUNIT),
    EPL_MOBILE = mean(EPL_MOBILE),
    EPL_CROWD = mean(EPL_CROWD),
    EPL_NOVEH = mean(EPL_NOVEH),
    EPL_GROUPQ = mean(EPL_GROUPQ),
    SPL_THEME4 = mean(SPL_THEME4),
    RPL_THEME4 = mean(RPL_THEME4)
  ) %>%
  ungroup()

# Merge county SVI with confirmed cases of Covid-19
SVI2_top_10_counties_cases <- merge(SVI2_top_10_counties_cases, ca_cases_top_10, by = 'county_fips') %>%
  arrange(desc(confirmed_cases)) %>%
  select(-county_name, -state_name)

# Select columns for correlation analysis
SVI2_top_10_counties_cases_corr <- SVI2_top_10_counties_cases[c('confirmed_cases','EPL_UNEMP','EPL_HBURD','EPL_NOHSDP','EPL_UNINSUR','SPL_THEME1','RPL_THEME1','EPL_AGE65','EPL_AGE17','EPL_DISABL','EPL_SNGPNT','EPL_LIMENG','SPL_THEME2','RPL_THEME2','EPL_MINRTY','SPL_THEME3','RPL_THEME3','EPL_MUNIT','EPL_MOBILE','EPL_CROWD','EPL_NOVEH','EPL_GROUPQ','SPL_THEME4','RPL_THEME4')]

# Calculate correlation matrix
corr <- cor(SVI2_top_10_counties_cases_corr)

corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.4,mar = c(0, 0, 2, 0), tl.cex = 0.4, tl.col = 'black', tl.srt = 45)

xkabledply(corr)
```


*Correlation matrix for ten counties with most deaths along different themes**

```{r}

# Select the rows in the SVI dataframe that contain the ten counties with most confirmed cases.
SVI2_top_10_counties_deaths <- SVI2[SVI2$county_fips %in% top_10_deaths_fips_list, ]

# Group by county, COUNTY, and STATE, and calculate the mean values for selected indicators.
SVI2_top_10_counties_deaths <- SVI2_top_10_counties_deaths %>%
  group_by(county_fips, COUNTY, STATE) %>%
  summarise(
    EPL_UNEMP = mean(EPL_UNEMP),
    EPL_HBURD = mean(EPL_HBURD),
    EPL_NOHSDP = mean(EPL_NOHSDP),
    EPL_UNINSUR = mean(EPL_UNINSUR),
    SPL_THEME1 = mean(SPL_THEME1),
    RPL_THEME1 = mean(RPL_THEME1),
    EPL_AGE65 = mean(EPL_AGE65),
    EPL_AGE17 = mean(EPL_AGE17),
    EPL_DISABL = mean(EPL_DISABL),
    EPL_SNGPNT = mean(EPL_SNGPNT),
    EPL_LIMENG = mean(EPL_LIMENG),
    SPL_THEME2 = mean(SPL_THEME2),
    RPL_THEME2 = mean(RPL_THEME2),
    EPL_MINRTY = mean(EPL_MINRTY),
    SPL_THEME3 = mean(SPL_THEME3),
    RPL_THEME3 = mean(RPL_THEME3),
    EPL_MUNIT = mean(EPL_MUNIT),
    EPL_MOBILE = mean(EPL_MOBILE),
    EPL_CROWD = mean(EPL_CROWD),
    EPL_NOVEH = mean(EPL_NOVEH),
    EPL_GROUPQ = mean(EPL_GROUPQ),
    SPL_THEME4 = mean(SPL_THEME4),
    RPL_THEME4 = mean(RPL_THEME4)
  ) %>%
  ungroup()

# Merge county SVI with confirmed cases of Covid-19
SVI2_top_10_counties_deaths <- merge(SVI2_top_10_counties_deaths, ca_deaths_top_10, by = 'county_fips') %>%
  arrange(desc(deaths)) %>%
  select(-county_name, -state_name)

# Select columns for correlation analysis
SVI2_top_10_counties_deaths_corr <- SVI2_top_10_counties_deaths[c('deaths','EPL_UNEMP','EPL_HBURD','EPL_NOHSDP','EPL_UNINSUR','SPL_THEME1','RPL_THEME1','EPL_AGE65','EPL_AGE17','EPL_DISABL','EPL_SNGPNT','EPL_LIMENG','SPL_THEME2','RPL_THEME2','EPL_MINRTY','SPL_THEME3','RPL_THEME3','EPL_MUNIT','EPL_MOBILE','EPL_CROWD','EPL_NOVEH','EPL_GROUPQ','SPL_THEME4','RPL_THEME4')]

# Calculate correlation matrix
corr <- cor(SVI2_top_10_counties_deaths_corr)

corrplot(corr, is.corr = TRUE, method = 'color', na.label='NA',col = colorRampPalette(c("white", "darkblue"))(100), addCoef.col = 'black', number.cex = 0.4,mar = c(0, 0, 2, 0), tl.cex = 0.4, tl.col = 'black', tl.srt = 45)

xkabledply(corr)
```





**Initial MLR for confirmed cases**
```{r}
library(caret)
# Preparing the independent variable (X) and the dependent variable (y)
y <- svi_all_counties_cases$confirmed_cases
#X <- svi_all_counties_cases[, c('EP_POV150','EP_LIMENG','EP_AGE65', 'EP_MINRTY',	'EP_UNEMP',	'EP_NOHSDP','EP_AGE17','EP_DISABL','EP_SNGPNT','EP_MUNIT','EP_MOBILE','EP_CROWD','EP_NOVEH','EP_GROUPQ')]
X <- svi_all_counties_cases[, c('RPL_THEMES','EP_NOHSDP', 'EP_AGE17', 'EP_MUNIT','EP_CROWD','EP_GROUPQ')]


# Splitting the data into train and test data
set.seed(42)  # For reproducibility
train_indices <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_indices, ]
X_test <- X[-train_indices, ]
y_train <- y[train_indices]
y_test <- y[-train_indices]

# Initiate the model
lm_model <- lm(y_train ~ ., data = cbind(y_train, X_train))

# Predictions on the test set
y_predict <- predict(lm_model, newdata = X_test)
summary(lm_model)
# Score (R-squared)
print(summary(lm_model)$r.squared)

# Coefficients
confirmed_cases_all_counties_coef <- data.frame(Coef = coef(lm_model)[-1], Features = names(coef(lm_model)[-1]))
confirmed_cases_all_counties_coef <- confirmed_cases_all_counties_coef[order(-confirmed_cases_all_counties_coef$Coef), ]

```

**Initial MLR for deaths**
```{r}
library(caret)
# Preparing the independent variable (X) and the dependent variable (y)
y <- svi_all_counties_deaths$deaths
#X <- svi_all_counties_cases[, c('EP_POV150','EP_LIMENG','EP_AGE65', 'EP_MINRTY',	'EP_UNEMP',	'EP_NOHSDP','EP_AGE17','EP_DISABL','EP_SNGPNT','EP_MUNIT','EP_MOBILE','EP_CROWD','EP_NOVEH','EP_GROUPQ')]
X <- svi_all_counties_deaths[, c('RPL_THEMES','EP_NOHSDP', 'EP_AGE17', 'EP_MUNIT','EP_CROWD','EP_GROUPQ')]


# Splitting the data into train and test data
set.seed(42)  # For reproducibility
train_indices <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_indices, ]
X_test <- X[-train_indices, ]
y_train <- y[train_indices]
y_test <- y[-train_indices]

# Initiate the model
lm_model <- lm(y_train ~ ., data = cbind(y_train, X_train))

# Predictions on the test set
y_predict <- predict(lm_model, newdata = X_test)
summary(lm_model)
# Score (R-squared)
print(summary(lm_model)$r.squared)

# Coefficients
deaths_all_counties_coef <- data.frame(Coef = coef(lm_model)[-1], Features = names(coef(lm_model)[-1]))
deaths_all_counties_coef <- deaths_all_counties_coef[order(-deaths_all_counties_coef$Coef), ]

```


**MLR with SVI2**
```{r}
library(caret)
# Preparing the independent variable (X) and the dependent variable (y)
y <- SVI2_all_counties_cases$confirmed_cases
X <- SVI2_all_counties_cases[,c('RPL_THEME1','RPL_THEME2','RPL_THEME3','RPL_THEME4')]


# Splitting the data into train and test data
set.seed(42)  # For reproducibility
train_indices <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_indices, ]
X_test <- X[-train_indices, ]
y_train <- y[train_indices]
y_test <- y[-train_indices]

# Initiate the model
lm_model <- lm(y_train ~ ., data = cbind(y_train, X_train))

# Predictions on the test set
y_predict <- predict(lm_model, newdata = X_test)
summary(lm_model)

# Score (R-squared)
print(summary(lm_model)$r.squared)

# Coefficients
confirmed_cases_all_counties_coef <- data.frame(Coef = coef(lm_model)[-1], Features = names(coef(lm_model)[-1]))
confirmed_cases_all_counties_coef <- confirmed_cases_all_counties_coef[order(-confirmed_cases_all_counties_coef$Coef), ]

```

**MLR with SVI2 for deaths**
```{r}
library(caret)
# Preparing the independent variable (X) and the dependent variable (y)
y <- SVI2_all_counties_deaths$deaths
X <- SVI2_all_counties_deaths[,c('RPL_THEME1','RPL_THEME2','RPL_THEME3','RPL_THEME4')]


# Splitting the data into train and test data
set.seed(42)  # For reproducibility
train_indices <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_indices, ]
X_test <- X[-train_indices, ]
y_train <- y[train_indices]
y_test <- y[-train_indices]

# Initiate the model
lm_model <- lm(y_train ~ ., data = cbind(y_train, X_train))

# Predictions on the test set
y_predict <- predict(lm_model, newdata = X_test)
summary(lm_model)

# Score (R-squared)
print(summary(lm_model)$r.squared)

# Coefficients
deaths_all_counties_coef <- data.frame(Coef = coef(lm_model)[-1], Features = names(coef(lm_model)[-1]))
deaths_all_counties_coef <- deaths_all_counties_coef[order(-deaths_all_counties_coef$Coef), ]

```


